Ornaments as indicators of social changes and cultural practices in northeast Taiwan before and after European colonial period

### 2019-12-03 analysis/paper/paper.Rmd
revision based on reviewer comments

One of the most commonly traded types of object in this region were personal ornaments such as glass and stone beads [@Chen2007; @Li2014; @NMTH2005]. Personal adornments in the archaeological record are useful as signal of an individual's status [@Joyce2005; @Scaramelli2005]. For example, the consumption of stone beads in Southeast Asia during Iron Age is often associated with increasing social stratification or socio-political complexity [@Bellina2014; @Carter2016; @Francis2002; @Theunissen2000; @Kenoyer2000]. In this paper, I explore archaeological ornaments from Kiwulan spanning the pre-European contact period, the period of Spanish and Dutch presence, and the period of Chinese presence. I address the question of whether indirect colonial influences on the indigenous populations can be detected through the ornament assemblages. 
# Historical Background of complex exchange network during the late Iron Age and early historical period

@@

Chinese historical records from 1829, 1837, and 1852 during the Qing dynasty (1616-1911) contain some notes on the purposes of ornaments from Yilan [@Chen1963, pp. 228, 308; @Ke1993, pp. 11, 126; @Yao1996, pp. 77]. According to those records, indigenous people in Yilan wore ornaments in ceremonial contexts to display their wealth and status. Among those ornaments, fish-shaped necklaces made of metal threads had high value due to their delicacy and the materials invested in production. These were usually possessed by wealthy people. Other people wore agate beads or glass beads on their head or neck to participate in ceremonies. In 1895, at the beginning of Japanese colonisation, an academic field survey for plains indigenous groups reported that golden necklaces were not used in Yilan at that time, but elderly people still used beads [@Ino1996, pp. 227-232]. In sum, we find multiple sources describing how ornaments represent high status or specialised social roles in indigenous communities in Yilan. Compared to the European period, there are fewer mentions of beads in Chinese period and the descriptions are limited to clothing, but generally confirm the role of beads as status markers. However, we must recognize some biases in historical records that might lead to misinterpretations of the use of ornaments. Inconsistent descriptions from different written sources and a lack of context in texts are the main problems need to be resolved by studying the ornaments in archaeological contexts.

Currently the ornaments found in northeast Taiwan in early historical period around 400 hundred years ago, including glass beads, stone beads, and metal ornaments, were considered as imported materials from other regions due to a lack of evidence of beadmaking waste, metal working, or accessible local raw materials. The chemical composition analysis for glass beads shows high content of lead, coupled with the winding/folding technique, suggest a Chinese beadmaking tradition [@Cheng2008; @Gan2006; @Wang2018]. Although there is wide varity of metal ornaments such as bells, bracelets, rings, and pendants, the common compositions of metal ornaments are brass and copper with a small number made from lead and tin that indicate multiple origins mostly likely from Southeast Asia [@Chen]. There is no clear evidence shows the European origin of beads, however, the large amount of golden beads at Kiwulan might be related to the colonial presence of the Europeans through economic activities because similar beads were found at Luzon, northern Philippines, a part of the trading route of the Spanish between 16-19th century [@Wang2007]. Northeast Taiwan from the late Iron age to early historical period was involved in complex exchange network both in a regional and global scale.

@@

This study focuses on `r sam_post_hole` ornaments from 40 sampling squares located at the dwelling place of Kiwulan, indicated by aliened postholes with in-situ posts (Figure \@ref(fig:KWL-sam-area)). These postholes were found aligned in a north-south direction in some intervals with construction marks, which were interpreted as the remains of silt house structures. At the north margin of the dwelling place were burials that are mostly oriented in an east-west direction across the sampling area. There are 30 burials in the sampling area account for one third of the total number of burials at Kiwulan. Ornaments are commonly used as grave goods in burials, and the total number including different ornament classes are 3,173. The use of ornaments as body adornments in burials will be discussed to understand the functions of ornaments, but not statistically analysed since the number of ornaments is highly affected by the presence of bead strings, which sometimes contains thousands of beads in an individual burial [@Chen2007]. In addition to ornaments from the dwelling place and burial, there are 27 ornaments found in middens, which to some extent supports that ornaments were treated as high value items at Kiwulan because of small number in the midden contexts compared to high frequency in burial contexts. It is important to analyse the detailed distribution of ornaments from the dwelling place as it can provide key information concerning social inequality based on uneven access to the high-value items, and changes in the spatial pattern of ornament across different time periods.

The ornaments found in the Figure \@ref(fig:plot-all-ornaments-count) shows that the most common ornament are golden beads (n = `r golden_bead_n`), followed by glass beads (n =  `r glass_bead_n`), carnelian beads (n = `r agate_bead_n`), metal rings (n = `r metal_ring_n`), and metal bells (n = `r metal_bell_n`, see also Figure \@ref(fig:ornaments-image)). The variety of subtypes and their individual frequencies are summarized in Table \@ref(tab:tidy-data-for-table).

@@

The distribution of frequencies for subtypes in each major class are presented in Figure \@ref(fig:plot-altogether). Spearman's correlation test shows that there is no significant relationship between diversity of subtypes and sample size (S = `r spearman_result_stat`, rho = `r spearman_result_rho`, p = $`r spearman_result_p_value`$). This indicates that the increases in diversity can be explained by the effects of culture interaction instead of effects of sample size. Carnelian beads and metal rings have greater quantity and variety of shapes compared to copper bell and glass beads during the European period. The greater varieties for carnelian beads and metal rings might indicate multiple origins due to participation in global trade networks stimulated by the European presence. In contrast, copper bells and glass beads have less variety, but glass beads have a larger number because they were usually found in clusters that indicate their original use as strings or necklaces. Copper bells are typically >2 cm long with a wide variety of human faces as a motif, while glass beads are usually <0.5 cm long Indo-Pacific beads, widespread in Southeast Asian sites since 300 BC [@Francis2002].

@@

An indirect colonial influence is indicated at Kiwulan by a greater diversity of ornaments types and materials during the European period. The origin of ornaments

This likely results from participation in large scale exchange networks that stimulated the circulation of different ornament classes. However, the frequency of overall ornaments and each subtype declines significantly during the Chinese period. This may be due to a smaller scale of trading network, the overall decline of the indigenous population in Yilan, and adopting the practices of Han Chinese who did not use these kinds of beads in their dress. Since the end of the 18th century, many indigenous people moved southwards to Hualien due to population pressure caused by Han Chinese immigrants [@Chen2007].

The explanations for the increase of ornaments during the European period could be

should be related to participation in global trade networks is not clarified.

On the contrary, a more economical interpretation could be that they are used to vindicate the local customs and resist against European intrusion, rather than the opposite, as the author does. Without using foreign elements (something possible in Kiwulan, where 300+ tobacco pipes have been found, for example) it is difficult to see the influence of global networks in any particular place.

Spatial patterns of ornaments shows that their distribution was clustered during the pre-European and European period. Testing for spatial patterns indicates that these clusters are non-random, and are most highly concentrated during the European period. Because ethnographic evidence indicates that trade ornaments were prestige goods in prehistoric Northeastern Taiwan, this spatial clustering may indicate that a degree of social inequality was present before European contact and then it was reinforced and amplified during the European period. A further indicator of increasing social inequality is a burial dated to the 17th century that included 60 golden beads, well above the average of 2-3 pieces in the pre-European period [@Chen2007; @Cheng2008].

How might these results fit into a bigger picture of social change at periphery of colonial systems? We may get some insight into the general pathways that led to social inequality in northeastern Taiwan by considering how people have achieved and maintained power in a wide variety of societies [@Drennan2010; @Feinman2000; @Ames2010; @Bowles2010]. For example, the corporate/network model proposed by @Feinman2000 expands traditional hierarchical complexity to provide a comparative basis for distinct strategies for power. In the network mode, inequality develops when individuals accumulate wealth through their individual networks and people use their wealth to attract factions, control resources, and monopolize trade networks. In contrast, the corporate mode stresses shared power across different groups and sectors, integrative ceremonies and rituals, and large cooperative labour tasks [@Feinman2000; @Siegel1999].

Yilan social organisation went from corporate mode before the European arrival, then transitioned to network mode during the European presence, then back to corporate mode in the Chinese period. The small number of ornaments, and less concentrated distribution during the pre-European period is consistent with shared power and wealth of the corporate mode. The long-distance trade network introduced by Europeans resulted in the appearance of a network mode due to competition among ambitious individuals for prestige, wealth, or power through collecting trade goods [@Brumfiel1994; @Clark1994]. Because of the weak direct control from the European colonizers in northeast Taiwan, local leaders may have had the flexibility to manipulate European colonial images, expand personal power, and monopolize the high-value trade goods [@Kang2012]. During the Chinese period, the decreasing frequency and diversity of ornaments show a decline in the production, use and discard of ornaments at Kiwulan. This reflects the smaller scale of exchange networks, with limited sources of ornaments, and limited opportunities for individuals to accumulate wealth.

>>>>

One of the most commonly traded types of object in this region were personal ornaments such as glass and stone beads [@Chen2007; @Li2014; @NMTH2005]. Personal adornments in the archaeological record are useful as signal of an individual's status [@Joyce2005; @Scaramelli2005]. For example, the consumption of stone beads in Southeast Asia during Iron Age is often associated with increasing social stratification or socio-political complexity [@Bellina2014; @Carter2016; @Francis2002; @Theunissen2000; @Kenoyer2000]. In this paper, I explore archaeological ornaments from Kiwulan spanning the pre-European contact period, the period of Spanish and Dutch presence, and the period of Chinese presence. I address the question of whether indirect colonial influences on the indigenous populations can be detected through the ornament assemblages.

# Historical Background of complex exchange network during the late Iron Age and early historical period

@@

Chinese historical records from 1829, 1837, and 1852 during the Qing dynasty (1616-1911) contain some notes on the purposes of ornaments from Yilan [@Chen1963, pp. 228, 308; @Ke1993, pp. 11, 126; @Yao1996, pp. 77]. According to those records, indigenous people in Yilan wore ornaments in ceremonial contexts to display their wealth and status. Among those ornaments, fish-shaped necklaces made of metal threads had high value due to their delicacy and the materials invested in production. These were usually possessed by wealthy people. Other people wore agate beads or glass beads on their head or neck to participate in ceremonies. In 1895, at the beginning of Japanese colonisation, an academic field survey for plains indigenous groups reported that golden necklaces were not used in Yilan at that time, but elderly people still used beads [@Ino1996, pp. 227-232]. In sum, we find multiple sources describing how ornaments represent high status or specialised social roles in indigenous communities in Yilan. Compared to the European period, there are fewer mentions of beads in Chinese period and the descriptions are limited to clothing, but generally confirm the role of beads as status markers. However, we must recognize the inherent biases in historical records and fragmented nature of texts selected by written record keepers that might lead to misinterpretations of artifacts [@Galloway2006]. Inconsistent and incomplete descriptions from different written sources lacking contexts are the problems need to be compared and examined by studying the ornaments in archaeological contexts.

Currently the ornaments found in northeast Taiwan in early historical period around 400 hundred years ago, including glass beads, stone beads, and metal ornaments, were considered as imported materials from other regions due to a lack of evidence of beadmaking waste, metal working, or accessible local raw materials. The chemical composition analysis for glass beads shows high content of lead, coupled with the winding/folding technique, suggest a Chinese beadmaking tradition [@Cheng2008; @Gan2006; @Wang2018]. Although there is wide varity of metal ornaments such as bells, bracelets, rings, and pendants, the common compositions of metal ornaments are brass and copper with a small number made from lead and tin that indicate multiple origins mostly likely from Southeast Asia [@Chen2011]. There is no clear evidence shows the European origin of beads, however, the large amount of golden beads at Kiwulan might be related to the colonial presence of the Europeans through economic activities because similar beads were found at Luzon, northern Philippines, a part of the trading route of the Spanish between 16-19th century [@Wang2007]. Northeast Taiwan from the late Iron age to early historical period was involved in complex exchange network both in a regional and global scale.

@@

This study focuses on `r sam_post_hole` ornaments from 40 sampling squares located at the dwelling place of Kiwulan, indicated by aliened postholes with in-situ posts (Figure \@ref(fig:KWL-sam-area)). These postholes were found aligned in a north-south direction in some intervals with construction marks, which were interpreted as the remains of silt house structures. At the north margin of the dwelling place were burials that are mostly oriented in an east-west direction across the sampling area. There are 30 burials in the sampling area account for one third of the total number of burials at Kiwulan. Ornaments are commonly used as grave goods in burials, and the total number including different ornament classes are 3,173. The use of ornaments as body adornments in burials will be discussed to understand the function of ornaments, but not statistically analysed since the amount of ornaments is highly affected by the presence of bead strands or patterned bands, which sometimes contains thousands of beads in an individual burial [@Chen2007]. In addition to ornaments from the dwelling place and burial, there are 27 ornaments found in middens, which to some extent supports the argument that ornaments were treated as high value items at Kiwulan based on a few finds in the midden contexts compared to high frequency in burial contexts. It is important to analyse the detailed distribution of ornaments from the dwelling place across different time periods as it can provide key information concerning social inequality indicated by uneven access to high-value item in relation to the foreign influences.

The ornaments found in the Figure \@ref(fig:plot-all-ornaments-count) shows that the most common ornament are golden beads (n = `r golden_bead_n`), followed by glass beads (n =  `r glass_bead_n`), carnelian beads (n = `r agate_bead_n`), metal rings (n = `r metal_ring_n`), and metal bells (n = `r metal_bell_n`, see also Figure \@ref(fig:ornaments-image)). The variety of subtypes and their individual frequencies are summarized in Table \@ref(tab:tidy-data-for-table). The ornaments found in the dwelling area were assumed as body adornments or a part of clothing in their daily life according to their use in the burial contexts.

@@

The distribution of frequencies for subtypes in each major class are presented in Figure \@ref(fig:plot-altogether). Spearman's correlation test shows that there is no significant relationship between diversity of subtypes and sample size (S = `r spearman_result_stat`, rho = `r spearman_result_rho`, p = $`r spearman_result_p_value`$). This indicates that the increases in diversity can be explained by the effects of culture interaction instead of effects of sample size. Carnelian beads and metal rings have greater quantity and variety of shapes compared to copper bell and glass beads during the European period. The greater varieties for carnelian beads and metal rings might indicate multiple origins due to participation in large scale trade networks stimulated by the European presence. In contrast, copper bells and glass beads have less variety that the bells are typically >2 cm long with a wide variation of human faces as a motif, while glass beads usually <0.5 cm long Indo-Pacific beads, widespread in Southeast Asian sites since 300 BC [@Francis2002].

@@

An indirect colonial influence is indicated at Kiwulan by a greater diversity of ornaments types and materials during the European period. Northeast Taiwan, Yilan, was involved in complex trading networks both in regional scale with other indigenous groups and Chinese merchants, and global scale with the European including the Dutch and the Spanish. The origin of ornaments might be multiple places mostly in Southeast Asia and introduced into northeast Taiwan by Chinese merchants before 17th century. Later, the trade activities became more frequent and intense in the 17th century that might be related to the colonial involvement of the Europeans based on the similar styles of golden beads found in Luzon that was once the colony of the Spanish during this period. In addition to ornaments, stonewares at Kiwulan that were also found on a Spanish shipwreck and Southeast Asia in relation to the Dutch and the Spanish trading routes, and the tobacco use were also believed introduced to Kiwulan by the Europeans [@Wang2007; @Chen2007]. The greater diversity and quantity of ornaments likely results from participation in large scale exchange networks that stimulated the circulation of different ornament classes. However, the frequency of overall ornaments and each subtype declines significantly during the Chinese period in the early 19th century. This may be due to a smaller scale of trading network, the overall decline of the indigenous population in Yilan, and adopting the practices of Han Chinese who did not use these kinds of beads in their dress. Since the end of the 18th century, many indigenous people moved southwards to Hualien due to population pressure caused by Han Chinese immigrants [@Chen2007]. Although historical records indicate that trade ornaments were prestige goods in prehistoric northeast Taiwan, the information might be biased that needs to be compared with archaeological evidence.

The archaeological contexts show that ornaments were rarely found in middens and mostly preserved well as body adornments in burial contexts serving as grave goods. This could reflect ornaments were viewed as valuable objects such as status indicators and their distribution across the dwelling area would help to understand the social relation reflected by people who possessed them, and the reaction of the society as a whole to the colonial presence in the 17th century. Spatial patterns of ornaments shows that their distribution was clustered during the pre-European and European period. Testing for spatial patterns indicates that these clusters are non-random, and are most highly concentrated during the European period. The spatial clustering during the pre-European and European period shows an increase in use of ornaments in the daily life at Kiwulan. It may indicate that a degree of social inequality based on uneven distribution was present before European contact and then it was reinforced and amplified during the European period. A further indicator of increasing social inequality is a burial dated to the 17th century that included 60 golden beads, well above the average of 2-3 pieces in the pre-European period [@Chen2007; @Cheng2008].

How might these results fit into a bigger picture of social change at periphery of colonial systems? We may get some insight into the general pathways that led to social inequality in northeastern Taiwan by considering how people have achieved and maintained power in a wide variety of societies [@Drennan2010; @Feinman2000; @Ames2010; @Bowles2010]. For example, the corporate/network model proposed by @Feinman2000 expands traditional hierarchical complexity to provide a comparative basis for distinct strategies for power. In the network mode, inequality develops when individuals accumulate wealth through their individual networks and people use their wealth to attract factions, control resources, and monopolize trade networks. In contrast, the corporate mode stresses shared power across different groups and sectors, integrative ceremonies and rituals, and large cooperative labour tasks [@Feinman2000; @Siegel1999]. This provides an interpretation that Yilan social organisation went from corporate mode before the European arrival, then transitioned to network mode during the European presence. The small number of ornaments, and less concentrated distribution during the pre-European period is consistent with shared power and wealth of the corporate mode. The long-distance trade network introduced by Europeans resulted in the appearance of a network mode due to competition among ambitious individuals for prestige, wealth, or power through collecting trade goods [@Brumfiel1994; @Clark1994]. Because of the weak direct control from the European colonizers in northeast Taiwan, local leaders may have had the flexibility to manipulate European colonial images, expand personal power, and monopolize the high-value trade goods [@Kang2012].

Another scenario could be that the increasing number and concentrated spatial pattern of ornaments present a form of resistance to European intrusion by using them to show social identity and emphasize the local custom had existed before European contact. Resistance could be presented in many forms, for example, @Rubertone discussed a case study in the North America that a large amount of native-made shell beads supposed to be tributes for the Europeans were found in native burial contexts. Similarly, the frequent presence of ornaments across the dwelling area not only exclusive in burial contexts suggests ornaments might be intentionally displayed in daily life to reinforce the cultural tradition. During the Chinese period, the decreasing frequency and diversity of ornaments show a decline in the production, use and discard of ornaments at Kiwulan. This reflects the smaller scale of exchange networks, with limited sources of ornaments, and limited opportunities for individuals to accumulate wealth. It is also possible that some ornaments were treated as heirlooms that passed down from one generation to another generation. However, the spacial pattern can also reflects the trend of distribution even they were heirlooms and this might suggest that the ornaments found in Chinese period might come from earlier period and continuously used till then. Further analysis about the source for each type of ornaments would be helpful to understand their origin and use context.
